---
title: "A CLIMBER Meta-Analysis 2024"
output: html_document
date: "2024-07-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setting up

```{r, echo=FALSE, message = FALSE}
# Libraries
library(metafor)
library(readxl)
library(devtools)
library(meta)
library(esvis)
library(dmetar)
library(tidyverse)
library(dplyr)
library(grid)
library(stringr)

# Number of digits for measures
settings.meta(digits = 1)
```

# Importing Datasets

```{r, echo=FALSE, message = FALSE}
## importing datasets from ODC-SCI (odc-sci.org)
## download data from odc-sci and place the csvs into the "data" folder
led_data <- read.csv("data/dataset_1082_1720635210.csv") #literature-extracted data (LED: odc-sci:1082) -- https://dx.doi.org/10.34945/F5DG6D
iad_data <- read.csv("data/dataset_1081_1720635248.csv") #individual animal data (IAD: odc-sci:1081) -- https://dx.doi.org/10.34945/F5J59P
```

# Calculations for Individual Animal Data

```{r, echo=FALSE, message = FALSE}
#calculating the pearson coefficient for IAD
iad_r_pre_post<-cor(iad_data$controlresult, iad_data$outcomeresult)

#Calculating effect size as standard mean difference for paired samples SMD
iad_data_agg <- iad_data%>%
  group_by(grouplabel, PMID)%>%
  summarise(
    n.treat = n(),
    n.control.adj = n.treat,
    mean_outcomeresult = mean(outcomeresult),
    mean_controlresult = mean(controlresult),
    controlresult_SD = sd(controlresult),
    outcomeresult_SD = sd(outcomeresult),
    SD_diff = sqrt(controlresult_SD^2+outcomeresult_SD^2-(2*iad_r_pre_post*
    controlresult_SD*outcomeresult_SD)))%>%# SD difference
    mutate(SD_diff = ifelse(SD_diff == 0, 0.4, SD_diff), # impute SD_diff 0 for study 16430371, since there is no variance.
    # SD within for the calculation of the paired SMD
    SD_within = SD_diff/sqrt(2*(1-iad_r_pre_post)),
    # paired SMD
    effect.size = (mean_outcomeresult-mean_controlresult)/SD_within,
    # Variance of the paired SMD
    Vd = (1/n.treat+(effect.size^2)/(2*n.treat))*(2*(1-iad_r_pre_post)), ##corrected from r_pre_post
    # SE of the paired SMD
    se.effect = sqrt(Vd))

# Add descriptive vars to the aggregate data set
 iad_data_summary <- iad_data %>%
      select(PMID, Animal, Strain, Sex_Grouped, Type_of_Injury, Injury_Segment_, Injury_Device, behavioraltest, outcomescore_max, Contusion_severity_group, grouplabel)%>%
  group_by(PMID, Animal, Strain, Sex_Grouped, Type_of_Injury, Injury_Segment_, Injury_Device, behavioraltest, outcomescore_max, Contusion_severity_group, grouplabel) %>%
  summarise()
 
iad_data_meta <- left_join(iad_data_summary, iad_data_agg)

## add sample size grouping now that n has been determined
iad_data_meta <- iad_data_meta %>%
    mutate( 
    samplesize = case_when( 
      between(n.treat, 0, 9) ~ "small (n<10)",
      between(n.treat, 10, 30) ~ "medium (10<=n<30)",
      between(n.treat, 30, 100)~ "large (n>30)"))

## removing 11 subjects from IAD where experimental grouping was unable to be determined 
iad_data_meta  <- iad_data_meta %>%
  filter(grouplabel != "*unknown*")

## removing group from study because there is only 1 subject
iad_data_meta  <- iad_data_meta %>%
  filter(grouplabel != "Study I: Drug")

```


### Calculations for Literature-Extracted Data

```{r, echo=FALSE, message = FALSE}
#Calculating pearson coefficient for LED
r_pre_post<-cor(led_data$controlresult, led_data$outcomeresult)

## Imputing variation based on the median coefficient of variation (CV)
## Median control CV
led_data_CV_control<-led_data%>%
  filter(control_outcomeerror != 0)%>%
  group_by(samplesize)%>%
  mutate(controlresult_SD = ifelse(SD_or_SEM == "SD", control_outcomeerror,
                                   control_outcomeerror*sqrt(n.control.adj)),
         controlresult_CV = controlresult_SD/controlresult)%>%
  filter(!is.infinite(controlresult_CV))%>%
  summarise(m_control_cv = median(controlresult_CV))

## Median outcome CV       
led_data_CV_outcome<-led_data%>%
  group_by(samplesize)%>%
  filter(outcome_outcomeerror != 0)%>%
  mutate(outcomeresult_SD = ifelse(SD_or_SEM == "SD", outcome_outcomeerror,
                                   outcome_outcomeerror*sqrt(n.treat)),
         outcomeresult_CV = outcomeresult_SD/outcomeresult)%>%
  filter(!is.infinite(outcomeresult_CV))%>%
  summarise(m_outcome_cv = median(outcomeresult_CV))

# Imputation of SD and calculation of the SMD as effect size, and SE considering paired samples.
led_data<-led_data%>%
  left_join(led_data_CV_control)%>%
  left_join(led_data_CV_outcome)

led_data<-led_data%>%
  mutate(
    ## create control SD var from SD or SEM
    controlresult_SD = ifelse(SD_or_SEM == "SD", control_outcomeerror,
                                   control_outcomeerror*sqrt(n.control.adj)),
    controlresult_SD = ifelse(is.na(SD_or_SEM), 0,controlresult_SD),
    ## create outcome SD var from SD or SEM
    outcomeresult_SD = ifelse(SD_or_SEM == "SD", outcome_outcomeerror,
                                   outcome_outcomeerror*sqrt(n.treat)),
    outcomeresult_SD = ifelse(is.na(SD_or_SEM), 0,outcomeresult_SD),
    ## Impute SD from CV
    controlresult_SD = ifelse(controlresult_SD == 0,
                                   controlresult*m_control_cv,controlresult_SD),
    outcomeresult_SD = ifelse(outcomeresult_SD == 0,
                                   outcomeresult*m_outcome_cv,outcomeresult_SD),
    SD_diff = sqrt(controlresult_SD^2+outcomeresult_SD^2-(2*r_pre_post*
    controlresult_SD*outcomeresult_SD)), ## SD difference
    #SD within
    SD_within = SD_diff/sqrt(2*(1-r_pre_post)),
    effect.size = (outcomeresult-controlresult)/SD_within,
    Vd = (1/n.treat+(effect.size^2)/(2*n.treat))*(2*(1-r_pre_post)),
    se.effect = sqrt(Vd)) ## SD within

```




```{r, echo=FALSE, message = FALSE}
## sorting for graphs
iad_data_meta <- iad_data_meta %>%
  arrange(PMID, grouplabel) %>%
  mutate(Contusion_severity_group = factor(Contusion_severity_group, levels = c("mild", "moderate", "severe")))

led_data<- led_data %>%
  arrange(PMID, grouplabel) %>%
  mutate(Contusion_severity_group = factor(Contusion_severity_group, levels = c("mild", "moderate", "severe")))

```

## Literature-Extracted Data Meta-analysis

```{r, echo=FALSE, message = FALSE}
############## meta-analysis for LED
led_m2<- metagen(effect.size, se.effect, data = led_data, sm = "SMD",
             n.e = n.treat,
             n.c = n.control.adj,
             studlab = grouplabel, method.tau = "REML", hakn = TRUE)

##subgroups
led_m2.pmid<- update(led_m2, byvar = PMID, tau.common = TRUE, bylab="Study PMID", order=order(led_m2$data$PMID))
led_m2.animal <- update(led_m2, byvar = Animal, tau.common = TRUE, bylab="Animal Type", order=order(led_m2$data$Animal))
led_m2.strain <- update(led_m2, byvar = Strain, tau.common = TRUE, bylab = "Animal Strain")
led_m2.sex <- update(led_m2, byvar = Sex_Grouped, tau.common = TRUE, bylab = "Animal Sex")
led_m2.levelinjury <- update(led_m2, byvar = Injury_Segment, tau.common = TRUE,bylab = "Level of injury") 
led_m2.severityinjury <- update(led_m2, byvar = Contusion_severity_group, tau.common = TRUE,bylab = "Injury severity") 
led_m2.samplesize <- update(led_m2, byvar = samplesize, tau.common = TRUE,bylab = "Sample Size") 

```

## Individual Animal Data Meta-analysis

```{r, echo=FALSE, message = FALSE}
############## meta-analysis for IAD
iad_m2<- metagen(effect.size, se.effect, data = iad_data_meta, sm = "SMD",
             n.e = n.treat,
             n.c = n.control.adj,
             studlab = grouplabel, method.tau = "REML", hakn = TRUE)      

##subgroups
iad_m2.pmid <- update(iad_m2, byvar = PMID, tau.common = TRUE, bylab="Study PMID", order=order(iad_m2$data$PMID))
iad_m2.animal <- update(iad_m2, byvar = Animal, tau.common = TRUE, bylab="Animal Type", order=order(iad_m2$data$Animal))
iad_m2.strain <- update(iad_m2, byvar = Strain, tau.common = TRUE, bylab = "Animal Strain")
iad_m2.sex <- update(iad_m2, byvar = Sex_Grouped, tau.common = TRUE, bylab = "Animal Sex")
iad_m2.levelinjury <- update(iad_m2, byvar = Injury_Segment_, tau.common = TRUE,bylab = "Level of injury")
iad_m2.severityinjury <- update(iad_m2, byvar = Contusion_severity_group, tau.common = TRUE,bylab = "Injury severity") 
iad_m2.samplesize <- update(iad_m2, byvar = samplesize, tau.common = TRUE,bylab = "Sample Size") 

```


```{r, echo=FALSE, message = FALSE}
## led data
mb3 <- metabind(led_m2.pmid)
mb4 <- metabind(led_m2.animal)
mb5 <- metabind(led_m2.animal, led_m2.sex, led_m2.strain)
mb6 <- metabind(led_m2.levelinjury, led_m2.samplesize)
mb7 <- metabind(led_m2.animal, led_m2.sex, led_m2.strain, led_m2.levelinjury, led_m2.samplesize, led_m2.severityinjury)

## iad data
mb14 <- metabind(iad_m2.pmid)
mb15 <- metabind(iad_m2.animal)
mb16 <- metabind(iad_m2.animal, iad_m2.sex, iad_m2.strain)
mb17 <- metabind(iad_m2.levelinjury, iad_m2.samplesize)
mb17.1 <- metabind(iad_m2.animal, iad_m2.sex, iad_m2.strain, iad_m2.levelinjury, iad_m2.samplesize,iad_m2.severityinjury)

## led & iad subgroup data
mb18 <- metabind(led_m2.animal, iad_m2.animal, name = c("Literature-Extracted Data", "Individual Animal Data"))
mb19 <- metabind(led_m2.strain, iad_m2.strain, name = c("Literature-Extracted Data", "Individual Animal Data"))
mb20 <- metabind(led_m2.sex, iad_m2.sex, name = c("Literature-Extracted Data", "Individual Animal Data"))
mb21 <- metabind(led_m2.severityinjury, iad_m2.severityinjury, name = c("Literature-Extracted Data", "Individual Animal Data"))
mb24 <- metabind(led_m2.levelinjury, iad_m2.levelinjury, name = c("Literature-Extracted Data", "Individual Animal Data"))
mb25 <- metabind(led_m2.samplesize, iad_m2.samplesize, name = c("Literature-Extracted Data", "Individual Animal Data"))
```

## Literature-Extracted Data Meta-Analysis Summary

```{r, echo=FALSE, fig.width = 9, fig.height = 10}

tiff("metagraphs_final_2024/led_summary.tiff", width = 8.5, height = 7.6, units = "in", res = 600)

forest_obj<-forest(mb3,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#ED6C6C",
       xlim = c(-20, 20),
       digits.addcols = 0,
       subgroup = FALSE,
       test.subgroup = FALSE,
       resid.hetstat = FALSE,
       just = "center",
       leftcols = c("studlab", "n.e"),
       leftlabs = c("Experimental Groups", "Sample Size"),
       addrow.subgroups = FALSE,
       common = FALSE,
       addrows.below.overall = 1)

grid.text("Random Effects Model: Literature-Extracted Data", y = unit(0.98, "npc"), gp = gpar(fontfamily = "Arial", fontsize = 13))

dev.off()

```

## Literature-Extracted Data Subgroups
```{r, echo=FALSE, fig.width = 9.5, fig.height = 6.5}
tiff("metagraphs_final_2024/led_subgroups.tiff", width = 8.5, height = 7, units = "in", res = 600)

forest(mb7,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#F1ACAC",
       xlim = c(-15, 15),
       digits.addcols = 0,
       colgap.forest.right = ".5cm",
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))

grid.text("Literature-Extracted Data Subgroups", y = unit(0.98, "npc"), gp = gpar(fontfamily = "Arial", fontsize = 13))

dev.off()

```




## Individual Animal Data Meta-Analysis Summary
```{r, echo=FALSE, fig.width = 10, fig.height = 8.2}

tiff("metagraphs_final_2024/iad_summary.tiff", width = 8.5, height = 7.6, units = "in", res = 600)

forest(mb14,
       lower.equi = iad_m2$lower.random, upper.equi = iad_m2$upper.random,
       fill.equi = "#FFD773",
       xlim = c(-20, 20),
       digits.addcols = 0,
       subgroup = FALSE,
       test.subgroup = FALSE,
       resid.hetstat = FALSE,
       colgap.left = "5mm",
       just = "center",
       leftcols = c("studlab", "n.e"),
       leftlabs = c("Experimental Groups", "Sample Size"),
       addrow.subgroups = FALSE,
       common = FALSE,
       addrows.below.overall = 1)

grid.text("Random Effects Model: Individual Animal Data", y = unit(0.98, "npc"), gp = gpar(fontfamily = "Arial", fontsize = 13))

dev.off()
```

## Individual Animal Data Subgroups
```{r, echo=FALSE, fig.width = 9, fig.height = 7}
tiff("metagraphs_final_2024/iad_subgroups.tiff", width = 8.5, height = 7, units = "in", res = 600)

forest(mb17.1,
       lower.equi = iad_m2$lower.random, upper.equi = iad_m2$upper.random,
       fill.equi = "#FFF88F",
       xlim = c(-15, 15),
       digits.addcols = 0,
       colgap.left = ".1mm",
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))

grid.text("Individual Animal Data Subgroups", y = unit(0.98, "npc"), gp = gpar(fontfamily = "Arial", fontsize = 13))

dev.off()
```

## Sex Published & Individual Animal Data
```{r, echo=FALSE, fig.width = 8, fig.height = 2.7}
tiff("metagraphs_final_2024/animal_sex.tiff", width = 8, height = 2.8, units = "in", res = 600)
forest(mb20,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#6A85A7",
       digits.addcols = 0,
       colgap.left = ".4cm",
       colgap.forest.right = ".5cm",
       xlim = c(-15, 15),
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))
dev.off()

```

## Animal Type Literature-Extracted Data & Individual Animal Data
```{r, echo=FALSE, fig.width = 8, fig.height = 2.7}
tiff("metagraphs_final_2024/animal_type.tiff", width = 8, height = 2.8, units = "in", res = 600)
forest(mb18,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#64C1D5",
       digits.addcols = 0,
       colgap.left = ".4cm",
       colgap.forest.right = ".5cm",
       xlim = c(-15, 15),
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))
dev.off()
```

## Animal Strain Literature-Extracted Data & Individual Animal Data
```{r, echo=FALSE, fig.width = 8, fig.height = 3}
tiff("metagraphs_final_2024/animal_strain.tiff", width = 8.2, height = 3, units = "in", res = 600)
forest(mb19,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#20AE9F",
       digits.addcols = 0,
       colgap.forest.right = ".5cm",
       xlim = c(-15, 15),
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))
dev.off()

```


## Injury Level Literature-Extracted Data & Individual Animal Data
```{r, echo=FALSE, fig.width = 7.8, fig.height = 3}
tiff("metagraphs_final_2024/injury_level.tiff", width = 7.8, height = 3.2, units = "in", res = 600)
forest(mb24,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#47C599",
       xlim = c(-15, 15),
       digits.addcols = 0,
       colgap.left = ".4cm",
       colgap.forest.right = ".5cm",
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))
dev.off()

```

## Sample Size Literature-Extracted Data & Individual Animal Data

```{r, echo=FALSE, fig.width = 8.5, fig.height = 3}
tiff("metagraphs_final_2024/samplesize.tiff", width = 8.5, height = 3.1, units = "in", res = 600)
forest(mb25,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#B9F0FC",
       xlim = c(-15, 15),
       digits.addcols = 0,
       colgap.forest.right = ".5cm",
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))
dev.off()
```


## Injury severity Literature-Extracted Data & Individual Animal Data
```{r, echo=FALSE, fig.width = 8.5, fig.height = 3}
tiff("metagraphs_final_2024/injury_severity.tiff", width = 8.2, height = 3.1, units = "in", res = 600)
forest(mb21,
       lower.equi = led_m2$lower.random, upper.equi = led_m2$upper.random,
       fill.equi = "#B6B3FF",
       xlim = c(-15, 15),
       digits.addcols = 0,
       colgap.forest.right = ".5cm",
       colgap.left = ".5cm",
       leftlabs = c("Variables", "Number of \nExperimental Groups", "Interaction \nP-value"))
dev.off()

default_parameters <- par()
default_parameters$family



```

